# Day 1 Block 4: Iterations with `map` {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


*Roughly 55 minutes. First the `map` toolkit, then the same tools applied to the
running example so that the three function calls from Block 3 become one pipeline.*

The refactored script still has three explicit calls, one per species. A fourth
species means a fourth call. Iteration removes the per-group calls entirely: the
function is applied to each group automatically.

## Part A: the `map` toolkit

`purrr::map(.x, .f)` applies the function `.f` to each element of `.x` and returns
a list. It is the tidyverse counterpart of base R's `lapply()`.

```{r}
#| eval: false
df_list <- list(a = 1:10, b = 11:20, c = 21:30)

purrr::map(df_list, mean)   # a list of three means
```

`map()` always returns a list, which is maximally general but often more than you
want. The typed variants return an atomic vector and *fail* if the result is not
of that type, which is a feature: it catches mistakes early.

| Function | Returns |
|---|---|
| `map()` | a list |
| `map_lgl()` | a logical vector |
| `map_int()` | an integer vector |
| `map_dbl()` | a double vector |
| `map_chr()` | a character vector |

```{r}
#| eval: false
purrr::map_dbl(df_list, mean)   # a named numeric vector, not a list
```

:::{.callout-note}

This type-safety is the reason to prefer `map_*()` over base `sapply()`.
`sapply()` guesses the output type and can return a vector, a matrix or a list
depending on the input, so a function that worked on Tuesday can silently return
a different shape on Wednesday. `map_dbl()` either returns a double vector or
errors. Predictability beats convenience.

:::

### Anonymous functions: `\(x)`, and the `~` you will still see

When the callback is not already a named function, write it inline. The modern
form is the R 4.1 shorthand `\(x)`:

```{r}
#| eval: false
purrr::map_dbl(df_list, \(x) sum(x) / length(x))
```

You will frequently meet the older purrr formula-lambda, `~ sum(.x) / length(.x)`,
where `.x` (or `.`) stands for the current element. It still works, but `\(x)` is
clearer because the argument has a real name, and it is ordinary R rather than
purrr-specific syntax. We use `\(x)` throughout.

Extra arguments can be passed either inside the lambda or as trailing arguments to
`map()`:

```{r}
#| eval: false
purrr::map_dbl(df_list, \(x) mean(x, na.rm = TRUE))   # inside the lambda
purrr::map_dbl(df_list, mean, na.rm = TRUE)           # forwarded by map()
```

### Returning a data frame: `map() |> list_rbind()`

To stack per-element tibbles into one tibble, run `map()` and then `list_rbind()`.

```{r}
#| eval: false
files <- list.files(here::here("data", "many_files"),
                    pattern = "\\.csv$", full.names = TRUE)

all_data <- files |>
  purrr::map(\(f) readr::read_csv(f, show_col_types = FALSE)) |>
  purrr::list_rbind()
```

:::{.callout-warning}

You will see `map_dfr()` and `map_dfc()` in older code and tutorials. These are
**superseded**: the current idiom is `map() |> list_rbind()` for row-binding and
`map() |> list_cbind()` for column-binding. They are clearer about the two
separate steps (apply, then combine) and are the forms the tidyverse now
recommends.

:::

### The base R family, in one paragraph

You will encounter `lapply()`, `sapply()`, `vapply()` and `apply()` in existing
code, and parallel iteration via `foreach` or `furrr`. The translation is direct:
`lapply(x, f)` is `map(x, f)`; `sapply(x, f)` is one of the typed `map_*()`
variants depending on the return type. Parallel back-ends such as `furrr::future_map()`
mirror `map()`'s interface for when a computation is genuinely slow and
embarrassingly parallel. We mention these so you recognise them; for new code,
`purrr` is the consistent default and parallelism is a later optimisation, applied
only after profiling shows it is needed.

### `map2` and `pmap`: iterating over more than one input

`map2(.x, .y, .f)` walks two collections in parallel, passing one element from
each to the function. `pmap(.l, .f)` generalises to any number of inputs, taking a
list (or a data frame, treating each column as an input).

```{r}
#| eval: false
sizes  <- c(2, 3, 4)
widths <- c(5, 6, 7)

purrr::map2_dbl(sizes, widths, \(s, w) s * w)   # 10 18 28
```

### `walk`: iteration for side effects

When you iterate to *do* something (write files, print plots) rather than to
*return* a value, use `walk()`/`walk2()`. They run the function for its side
effect and return their input invisibly, so the console is not flooded with output.

```{r}
#| eval: false
# Write each split of the data to its own file, silently
peng_split <- penguins_clean |> 
  dplyr::group_by(species) |> 
  dplyr::group_split()

names(peng_split) <- c("Adelie", "Chinstrap", "Gentoo")


purrr::walk2(
  peng_split, names(peng_split),
  \(data, nm) readr::write_csv(data, here::here("data", paste0(nm, ".csv")))
)

```

## Part B: iteration on the running example 

Recall the list-column from Block 2: `group_by(species) |> nest()` puts each
species' tibble into one cell of a `data` column.

```{r}
#| eval: false

species_data <- penguins_clean |>
  dplyr::group_by(species) |>
  tidyr::nest()

species_data
# # A tibble: 3 x 2
#   species   data
#   <fct>     <list>
# 1 Adelie    <tibble [146 x 7]>
# 2 Gentoo    <tibble [119 x 7]>
# 3 Chinstrap <tibble [68 x 7]>

```

:::{.callout-note}

We use the classic `group_by() |> nest()` rather than the newer
`nest(.by = species)`. Both give the same result; the explicit grouping step is
clearer and backward-compatible.

:::

### Applying a function to each cell

Used inside `mutate()`, `map()` applies a function to each cell of a list-column
and stores the result as a new list-column:

```{r}
#| eval: false
species_models <- species_data |>
  dplyr::mutate(
    model = purrr::map(data, \(d) lm(body_mass_g ~ flipper_length_mm, data = d)),
    coefs = purrr::map(model, \(m) broom::tidy(m, conf.int = TRUE)),
    n_obs = purrr::map_int(data, nrow)
  )

species_models
# # A tibble: 3 x 5
#   species   data              model  coefs            n_obs
#   <fct>     <list>            <list> <list>           <int>
# 1 Adelie    <tibble [146x7]>  <lm>   <tibble [2x7]>     146
# 2 Gentoo    <tibble [119x7]>  <lm>   <tibble [2x7]>     119
# 3 Chinstrap <tibble [68x7]>   <lm>   <tibble [2x7]>      68
```

Three things to notice. The `model` cell holds a fitted `lm` per species; the
`coefs` cell holds a tidied tibble per species; and `n_obs` is an ordinary integer
column, not a list-column, because `map_int()` returns an atomic vector. The typed
variants slot straight into a tibble.

### Unnesting back to a flat table

For a manuscript coefficient table, expand the `coefs` list-column back into rows.
`unnest()` is the dual of `nest()`: each cell's tibble becomes rows, with the
surrounding columns repeated.

```{r}
#| eval: false
coefficient_table <- species_models |>
  dplyr::ungroup() |>
  dplyr::select(species, coefs) |>
  tidyr::unnest(coefs)
```

### A real refactor: decompose the function

There is a tension. `fit_species_model()` filters *and* fits, but `nest()` has
already done the filtering. Calling it inside the iteration would filter twice.
That tension is information: the function is doing two jobs. Split them.

```{r}
#| eval: false
# R/functions/modelling_functions.R -- revised

#' Fit body mass ~ flipper length on pre-filtered data
#' @param data A data frame, assumed to be one species' worth.
#' @return A fitted lm object.
fit_model <- function(data) {
  lm(body_mass_g ~ flipper_length_mm, data = data)
}

#' Tidy a fitted lm with confidence intervals
#' @param model A fitted lm object.
#' @param conf_level Confidence level. Default 0.95.
#' @return A tibble of tidied coefficients.
tidy_model <- function(model, conf_level = 0.95) {
  broom::tidy(model, conf.int = TRUE, conf.level = conf_level)
}

#' Fit for one species (wrapper preserving the old interface)
#' @param data A data frame with a `species` column.
#' @param sp Character of length 1. Species to filter to.
#' @param conf_level Confidence level. Default 0.95.
fit_species_model <- function(data, sp, conf_level = 0.95) {
  if (!sp %in% unique(data$species)) {
    stop("Species '", sp, "' is not in the data. ",
         "Available species: ",
         paste(sort(unique(data$species)), collapse = ", "))
  }
  data |>
    dplyr::filter(species == sp) |>
    fit_model() |>
    tidy_model(conf_level = conf_level) |>
    dplyr::mutate(species = sp, .before = 1)
}
```

`fit_model()` and `tidy_model()` now exist independently and compose. The old
public interface survives as a wrapper, so existing callers and the tests from
Block 3 still pass with no edits. That is what a good refactor looks like:
behaviour preserved, internals improved, tests green.

### The iteration, rewritten

With single-responsibility functions, the iteration uses them directly and reads
cleanly:

```{r}
#| eval: false
species_models <- penguins_clean |>
  dplyr::group_by(species) |>
  tidyr::nest() |>
  dplyr::mutate(
    model = purrr::map(data, fit_model),
    coefs = purrr::map(model, tidy_model)
  )

coefficient_table <- species_models |>
  dplyr::ungroup() |>
  dplyr::select(species, coefs) |>
  tidyr::unnest(coefs)
```

This is the destination. The modelling logic that was three copy-pasted blocks in
Block 1, then three function calls in Block 3, is now one pipeline that applies the
same function to every group.

### Activity: iterate over a different grouping

:::{.callout-important}
## Task

The penguins data has an `island` column ("Biscoe", "Dream", "Torgersen"). Fit a
separate `body_mass_g ~ flipper_length_mm` model for each island and produce a
tidied coefficient table, using the nest pattern.

:::

:::{.callout-tip}
## Solution

```{r, eval = FALSE}
island_coefs <- penguins_clean |>
  dplyr::group_by(island) |>
  tidyr::nest() |>
  dplyr::mutate(
    model = purrr::map(data, fit_model),
    coefs = purrr::map(model, tidy_model)
  ) |>
  dplyr::ungroup() |>
  dplyr::select(island, coefs) |>
  tidyr::unnest(coefs)
```

`fit_model()` and `tidy_model()` are reused unchanged. They never knew they would
iterate over species; they work for islands too, because each has a single
responsibility and no hidden assumption about the grouping.

:::

### Extension exercise: plots in a list-column

:::{.callout-important}
## Task

A list-column can hold plots as readily as models. Using the nested data, add a
`plot` column made with `map()` that holds one scatter plot per species, then draw
them together with `patchwork::wrap_plots()`. Hint: build each plot inside
`\(d) ggplot(d, ...)`.

:::

:::{.callout-tip}
## Solution

```{r, eval = FALSE}
species_plots <- penguins_clean |>
  dplyr::group_by(species) |>
  tidyr::nest() |>
  dplyr::mutate(
    plot = purrr::map(data, \(d)
      ggplot2::ggplot(d, ggplot2::aes(flipper_length_mm, body_mass_g)) +
        ggplot2::geom_point(alpha = 0.6) +
        ggplot2::geom_smooth(method = "lm") +
        ggplot2::theme_minimal())
  )

patchwork::wrap_plots(species_plots$plot)
```

The same `nest |> mutate(map(...))` pattern that fits models also builds figures.
Once you see a cell as "any R object", models, plots and tables are all just
list-columns.

:::

:::{.callout-note}

**Where this leaves us.** Copy-paste is gone. One pipeline fits a model per group,
whatever the group is. But `fit_model()` still hardcodes the response and
predictor: it only ever fits body mass on flipper length. Block 5 removes that last
limitation.

:::
