
# SQL and Data.Table


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r workshop-setup}
#| include: false
knitr::opts_chunk$set(echo      = TRUE,
                      eval      = FALSE,
                      message   = FALSE,
                      warning   = FALSE,
                      fig.align = "center")
```

This chapter is both a teaching script for a two-hour session and a self-study
handout. It rests on one premise: correct syntax and a tool suggestion are cheap.
You can get them from documentation, a colleague, a forum answer, or a generated
snippet. What none of those sources can do is run a benchmark on *your* data on
*your* machine, or tell you whether a general recommendation holds for your file
and your most common query. Those answers require measurement, and measurement
requires understanding the mechanism well enough to design the right test.

The session therefore leads with the parts you cannot delegate. By the end you
should be able to do one thing reliably: decide between CSV and Parquet for a
given task, and read large data lazily so that only the result enters memory
rather than the whole file. Everything else in the session serves that single
capability.

It assumes familiarity with `dplyr` and the native pipe (`|>`). No prior
experience with `data.table`, Arrow, or DuckDB is assumed. Where `data.table` and
SQL appear, the aim is that you can *read* them, not write them fluently; writing
references are provided as take-home appendices.

:::{.callout-note}
**How to use this document.**

The numbered blocks below are written to be delivered live. Each exercise
contains complete, copyable code and a worked solution in a collapsible box, so
you can return afterwards and repeat it independently.

**Dependencies.** Install the following before the session:

    install.packages(c("readr", "data.table", "dtplyr", "arrow",
                       "duckdb", "dbplyr", "dplyr", "bench"))

**Data.** The test files are distributed in advance. You do **not** generate them
during the session. Confirm they load using the setup-check script below before
you arrive.
:::

## Running order {-}

| Block | Content | Time |
|---|---|---|
| 0 | Setup confirmation | ≈2 min |
| 1 | Framing: what you can delegate and what you must check | ≈3 min |
| 2 | Data formats: mechanisms on disk | ≈18 min |
| 3 | Lazy versus eager, and the `collect()` pattern | ≈12 min |
| 4 | Exercise 1: column selectivity and lazy memory, measured | ≈20 min |
| — | Break | ≈10 min |
| 5 | Query optimisation: filter early, select early | ≈12 min |
| 6 | Exercise 2: finding and fixing an inefficient query | ≈14 min |
| 7 | Reading the other syntaxes through `show_query()` | ≈12 min |
| 8 | Verifying code you did not write line by line | ≈10 min |
| 9 | Close: the one capability, and the checklist | ≈7 min |

:::{.callout-warning}
**If you run over, cut in this order.** Compress block 7 to a single worked
example and let the syntax reference survive as a take-home appendix. If still
over, drop principle two from block 5. Protect blocks 2, 3, and 4: the format
mechanisms, the lazy/eager distinction, and the measured exercise are the
take-home capability, and they are the least replaceable content in the session.
:::

---

## Setup: preparing and confirming the data {-}

### Instructor: build the files once, ahead of time {-}

Run this script once before distributing the files. It generates a one-million-row
synthetic survey dataset and writes it in three forms. Distribute the resulting
`bird_survey.csv`, `bird_survey.parquet`, and `bird_survey_partitioned/` so that
no participant regenerates a million rows live.

```{r generate-data}
#| eval: false
library(readr);   library(data.table); library(dtplyr)
library(arrow);   library(duckdb);     library(dbplyr)
library(dplyr);   library(bench)

set.seed(42)
n <- 1e6

species_names <- c(
  "Parus major", "Turdus merula", "Erithacus rubecula",
  "Fringilla coelebs", "Passer domesticus", "Columba palumbus",
  "Corvus corone", "Sturnus vulgaris", "Sylvia atricapilla",
  "Phylloscopus collybita"
)

survey_data <- tibble(
  observation_id      = 1:n,
  species             = sample(species_names, n, replace = TRUE),
  site_id             = sample(paste0("SITE_", sprintf("%04d", 1:500)),
                               n, replace = TRUE),
  habitat             = sample(c("woodland", "grassland", "wetland",
                                 "urban", "heathland"), n, replace = TRUE),
  year                = sample(2000:2023, n, replace = TRUE),
  month               = sample(1:12, n, replace = TRUE),
  day                 = sample(1:28, n, replace = TRUE),
  observer_id         = sample(paste0("OBS_", sprintf("%03d", 1:100)),
                               n, replace = TRUE),
  count               = rpois(n, lambda = 3),
  temperature_c       = rnorm(n, mean = 12, sd = 5),
  wind_speed_ms       = rgamma(n, shape = 2, rate = 0.5),
  precipitation_mm    = rgamma(n, shape = 0.5, rate = 1),
  visibility_m        = pmax(rnorm(n, mean = 5000, sd = 1500), 100),
  canopy_cover_pct    = runif(n, 0, 100),
  survey_duration_min = sample(c(5L, 10L, 15L, 20L), n, replace = TRUE)
)

write_csv(survey_data,     "bird_survey.csv")
write_parquet(survey_data, "bird_survey.parquet")
write_dataset(survey_data,
              path         = "bird_survey_partitioned",
              partitioning = "year")
```

### Participant: confirm the files load (run before the session) {-}

This prints package versions and the file sizes. Send the printed line back to the
instructor before the session so anyone whose installation failed can be helped in
advance. The likely failure points are the `arrow` and `duckdb` binary installs;
if either errors, reply rather than waiting for the day.

```{r setup-check}
#| eval: false
library(readr) 
library(arrow) 
library(duckdb)
library(dplyr)

stopifnot(
  file.exists("bird_survey.csv"),
  file.exists("bird_survey.parquet"),
  dir.exists("bird_survey_partitioned")
)

cat("R       ", as.character(getRversion()), "\n")
cat("arrow   ", as.character(packageVersion("arrow")),  "\n")
cat("duckdb  ", as.character(packageVersion("duckdb")), "\n")
cat("CSV:     ", round(file.size("bird_survey.csv")     / 1e6, 1), "MB\n")
cat("Parquet: ", round(file.size("bird_survey.parquet") / 1e6, 1), "MB\n")
```

:::{.callout-note}
**Instructor.** The CSV-versus-Parquet size ratio in the final two lines is the
first concrete illustration of columnar compression, before any benchmarking.
Draw attention to the ratio at the start of block 2.
:::

---

## 1. Framing: what you can delegate and what you must check (≈3 min) {-}

Generating syntactically correct code to read a Parquet file, or producing a
decision table that maps file size to a recommended tool, is routine. It is also
not where this session spends its time, because it is handled. Treat any code you
did not write and check line by line the same way, whether it came from a
colleague, a forum, your own past self, or a generated suggestion.

What no source of code can do for you is run a benchmark on your data on your
machine. It cannot tell you whether `fread` is four times or seven times faster
than `read_csv` on your specific file, whether the partition structure you chose
prunes effectively for your most common filter, or whether a general performance
claim survives contact with your dataset. Those answers require measurement, and
measurement requires understanding the mechanism well enough to design the right
test.

The reasoning and the checking are the content of this session. The syntax you
can look up.

---

## 2. Data formats: mechanisms on disk (≈18 min) {-}

Performance differences between tools are not arbitrary. They follow from how each
format stores data on disk. Understanding the mechanism is what lets you evaluate
a recommendation rather than accept it.

A note on a term used throughout: **I/O** is short for input/output, meaning the
work of reading from and writing to disk. Disk I/O is usually far slower than
computation in memory, so a format that reads fewer bytes from disk is often the
faster one.

### CSV: row-oriented text {-}

Open a CSV file in a text editor and you see exactly what R reads. Each line is
one observation; all columns for that observation appear on the same line,
separated by commas.

```
Row 1: observation_id, species, site_id, habitat, year, ..., all 15 columns
Row 2: observation_id, species, site_id, habitat, year, ..., all 15 columns
```

This **row-oriented** layout has one important consequence. To retrieve `species`
and `count` from row 1, the parser reads all 15 values, extracts 2, and discards
13. It does this for every row. Requesting fewer columns reduces what is stored in
R memory, but the parser still reads the full file from disk. Column selection is
a memory optimisation, not a speed optimisation, for CSV.

CSV also has no type system. Every value is text. The reader scans the first
thousand rows and infers whether a column holds numbers or characters. If unusual
values appear only after row one thousand, the inference may be wrong.

### Parquet: columnar binary {-}

Parquet uses a fundamentally different arrangement. Each column occupies a
separate contiguous block on disk:

```
[all 1M values of species] [all 1M values of habitat] [all 1M values of count] ...
[footer: column names, types, and min/max statistics per row group]
```

Requesting `species` and `count` touches only those two blocks. The other 13
columns are never read from disk. This is called **column projection**: reading
only the columns you ask for. Column selection therefore reduces both disk I/O and
memory for Parquet. This is the central structural difference from CSV, and it is
the take-home of the session.

The footer stores the minimum and maximum value of each column within each **row
group**, which is a horizontal batch of rows (a few tens of thousands by default).
A query engine can read the footer and skip any row group whose range cannot
satisfy a filter. If you filter `year == 2020` and a row group spans years 2000 to
2019, that group is skipped without reading its data. Parquet also stores column
types in the footer, so the reader never guesses types.

### Compression {-}

Parquet compresses each column block independently. The `arrow` default is Snappy:
fast to compress and decompress, with a moderate ratio. Gzip compresses more
aggressively but decompresses more slowly. Zstandard (zstd) typically matches or
improves on Snappy's ratio at comparable speed.

```{r compression-comparison}
#| eval: false
tibble(
  format  = c("CSV", "Snappy", "Gzip", "Zstd"),
  size_mb = c(
    file.size("bird_survey.csv"),
    file.size("bird_survey.parquet"),                       # default Snappy
    {
      write_parquet(read_parquet("bird_survey.parquet"),
                    "bird_survey_gzip.parquet", compression = "gzip")
      file.size("bird_survey_gzip.parquet")
    },
    {
      write_parquet(read_parquet("bird_survey.parquet"),
                    "bird_survey_zstd.parquet", compression = "zstd")
      file.size("bird_survey_zstd.parquet")
    }
  ) / 1e6
) |>
  mutate(ratio_vs_csv = round(size_mb[1] / size_mb, 1))
```

### Partitioned datasets and pruning {-}

A **partitioned dataset** is a directory of Parquet files organised by the values
of one or more columns: one subdirectory per year, for example. A query engine can
skip entire subdirectories when a filter excludes those values. This is
**partition pruning**, and it activates only when you filter on the partition key.

```{r partition-demo}
#| eval: false
# Filters on a non-partition column: every directory is scanned
open_dataset("bird_survey_partitioned") |>
  filter(habitat == "woodland") |>          # habitat is not the partition key
  summarise(total = sum(count)) |>
  collect()

# Filters on the partition key: only the matching directory is read
open_dataset("bird_survey_partitioned") |>
  filter(year == 2020) |>                   # year IS the partition key
  summarise(total = sum(count)) |>
  collect()
```

:::{.callout-note}
**Instructor (≈6 min).** Show the file-size table, then the
`bird_survey_partitioned/` directory: the `year=2000` ... `year=2023`
subdirectories make the partition key visible. Connect this to the code above:
filtering `year == 2020` opens only the `year=2020/` directory.
:::

:::{.callout-warning}
Partition pruning works only on the partition key. Filtering on any other column
still scans all partitions. Partition around the variable you filter most often.
For longitudinal monitoring data that is usually `year`; if you almost always
filter by region, partition by region instead.
:::

---

## 3. Lazy versus eager, and the `collect()` pattern (≈12 min) {-}

There are two ways to bring data into a query. **Eager** evaluation runs
immediately and returns a result you can see: `read_csv()` and `read_parquet()`
are eager, and they pull the whole file into memory. **Lazy** evaluation records
*what you intend to do* without running it, then executes only when you ask. The
boundary between the two is `collect()`. Before `collect()` you hold a query
description and nothing is in R; after `collect()` you hold a tibble.

```{r lazy-class-check}
#| eval: false
lazy_query <- open_dataset("bird_survey.parquet") |>
  filter(habitat == "woodland") |>
  summarise(total = sum(count))

class(lazy_query)         # a query description, NOT a tibble

lazy_query |> collect()   # now it runs, and returns a tibble
```

Why this matters for the take-home: a lazy read combined with Parquet means the
engine reads only the columns and row groups your query needs, and only the result
enters memory. If your file approaches the RAM you have, the lazy path is the only
one that fits, and a benchmark on a small test file will not reveal that. You have
to understand the mechanism to anticipate it.

Three tools share this lazy pattern. You write `dplyr` verbs to describe the
query, then call `collect()`.

- **`dtplyr`** runs the query on a `data.table` back-end. Wrap an in-memory table
  with `lazy_dt()`.
- **`arrow`** runs it on the Arrow engine over Parquet files, via `open_dataset()`.
- **`duckdb`** runs it inside an in-process database engine. "In-process" means
  the engine runs inside your R session, with no separate database server to
  install or connect to, unlike PostgreSQL or MySQL. You reach it through `tbl()`,
  using `DBI`, R's standard database interface.

```{r collect-pattern}
#| eval: false
# arrow back-end
open_dataset("bird_survey.parquet") |>
  filter(habitat == "woodland") |>
  group_by(species, year) |>
  summarise(total = sum(count), .groups = "drop") |>
  collect()                                            # execute -> tibble

# DuckDB back-end. Connect once, then expose the file as a view.
con <- dbConnect(duckdb())
dbExecute(con, "CREATE VIEW surveys_parquet AS
                SELECT * FROM read_parquet('bird_survey.parquet')")

tbl(con, "surveys_parquet") |>
  filter(habitat == "woodland") |>
  group_by(species, year) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  collect()                                            # execute -> tibble
```

:::{.callout-note}
**Instructor.** Spend two minutes on `collect()` explicitly. Participants who grasp
that everything before `collect()` is a description, not a result, will read the
Exercise 1 memory benchmark correctly. Note the DuckDB pattern: a file is exposed
with `CREATE VIEW ... read_parquet(...)`, then queried by name. We register
in-memory data frames differently, in block 8.
:::

---

## 4. Exercise 1: column selectivity and lazy memory, measured (≈20 min) {-}

:::{.callout-important}
Before running any code, write down your prediction for each benchmark. The
prediction is the exercise: comparing what you expected to what you measured is
where the understanding forms.
:::

:::{.callout-warning}
**Two things to know about the measurements.** `bench::mark()` reports `mem_alloc`
as memory allocated **on the R heap only**. Engines that do work outside R, such as
Arrow and DuckDB, may show deceptively small `mem_alloc` because the real
allocation happens in their own memory, not R's. Treat `mem_alloc` as indicative
for these engines and watch your system memory if memory pressure is the question.
Second, three to ten iterations give noisy figures. The lesson is the direction and
rough magnitude of an effect, not the exact ratio. Re-run a block and you will see
the numbers move.
:::

### Part A: does column selection reduce speed, or only memory? {-}

You are selecting 2 of 15 columns, roughly 13% of the data. If column selection
reduces disk I/O, you would expect a speed improvement. If it only reduces memory,
speed should be similar.

```{r ex1-part-a}
#| eval: false
bench::mark(
  csv_15_cols     = read_csv("bird_survey.csv", show_col_types = FALSE),
  csv_2_cols      = read_csv("bird_survey.csv",
                             col_select     = c(species, count),
                             show_col_types = FALSE),
  parquet_15_cols = read_parquet("bird_survey.parquet"),
  parquet_2_cols  = read_parquet("bird_survey.parquet",
                                 col_select = c(species, count)),
  iterations = 5,
  check      = FALSE
)[, c("expression", "median", "mem_alloc")]
```

**Questions.**

1. For CSV, does `median` time fall with column selection, or does only
   `mem_alloc` fall?
2. For Parquet, do both `median` and `mem_alloc` fall? What does that confirm about
   columnar storage?
3. Someone recommends `col_select` on a CSV file "for speed". Based on this
   benchmark, is that correct?

`r hide("Solution")`

For CSV, `mem_alloc` should fall substantially, from roughly 15 columns' worth to
2, but `median` time should change little. The parser still reads all 15 columns
from disk for every row and discards 13 before storing the result. Column selection
is a memory optimisation for row-oriented formats.

For Parquet, both `median` and `mem_alloc` should fall. The Arrow reader touches
only the two requested column blocks; the other 13 are never read from disk. This
is column projection, and the speed improvement reflects a genuine reduction in
disk I/O, not just less copying into R.

The recommendation is wrong for CSV and right for Parquet. The error is treating a
format-specific behaviour as a general rule.

`r unhide()`

### Part B: eager versus lazy memory {-}

```{r ex1-part-b}
#| eval: false
bench::mark(
  eager = read_parquet("bird_survey.parquet") |>
    filter(habitat == "woodland") |>
    summarise(total = sum(count)),

  lazy  = open_dataset("bird_survey.parquet") |>
    filter(habitat == "woodland") |>
    summarise(total = sum(count)) |>
    collect(),

  iterations = 5,
  check = FALSE
)[, c("expression", "median", "mem_alloc")]
```

**Questions.**

1. Which approach allocates more R-heap memory, and why?
2. The test file is about 90 MB. Suppose your real file is 40 GB on a 16 GB
   machine. Which approach can run at all? Could this benchmark have told you that?

`r hide("Solution")`

`eager` reads the entire file into R, then filters and summarises, so its
`mem_alloc` reflects the full file. `lazy` pushes the filter and the summary into
the Arrow engine and brings back only the one-row result, so its R-heap allocation
is small (subject to the off-heap caveat above).

For the 40 GB file, only the lazy approach can run, because the eager read would
require the whole file in 16 GB of RAM. The 90 MB benchmark cannot reveal this:
both approaches succeed at 90 MB. This is exactly the case where you must reason
from the mechanism rather than from a small-file measurement.

`r unhide()`

---

## 5. Query optimisation: filter early, select early (≈12 min) {-}

Query structure affects performance independently of the tool. A poorly structured
query on a fast engine can lose to a well-structured query on plain `dplyr`. Two
principles carry most of the benefit.

### Principle 1: filter early {-}

Applying a filter before an aggregation or join reduces the data flowing through
every later step. Apply it afterwards and the expensive operation runs on the full
dataset.

```{r filter-early}
#| eval: false
surveys <- read_csv("bird_survey.csv", show_col_types = FALSE)

bench::mark(
  filter_late  = surveys |>
    group_by(species, habitat, year) |>
    summarise(total = sum(count), .groups = "drop") |>
    filter(habitat == "woodland"),         # filter AFTER aggregating all habitats

  filter_early = surveys |>
    filter(habitat == "woodland") |>       # filter BEFORE aggregating
    group_by(species, year) |>
    summarise(total = sum(count), .groups = "drop"),

  iterations = 10,
  check = FALSE
)[, c("expression", "median", "mem_alloc")]
```

The point is the mechanism, not the precise ratio: `filter_late` aggregates all
five habitats and then discards four fifths of the result, while `filter_early`
discards them first. On a fast in-memory engine the gap can be modest and noisy, so
read the direction across re-runs rather than trusting a single figure. In SQL the
same idea is the distinction between `WHERE` (before grouping) and `HAVING` (after
grouping); use `WHERE` whenever the condition does not depend on an aggregate.

### Principle 2: select early {-}

Carrying unneeded columns through a pipeline costs memory and, for Parquet, disk
I/O. Select the minimum columns before any heavy step. On a lazy Parquet or DuckDB
source, an early `select()` becomes column projection at the point of reading.

```{r select-early}
#| eval: false
open_dataset("bird_survey.parquet") |>
  select(species, habitat, year, count) |>   # 4 columns read from disk, not 15
  filter(habitat == "woodland") |>
  group_by(species, year) |>
  summarise(total = sum(count), .groups = "drop") |>
  collect()
```

---

## 6. Exercise 2: finding and fixing an inefficient query (≈14 min) {-}

:::{.callout-important}
The query below has a structural problem. Identify it, rewrite the query, and
benchmark both versions to confirm the improvement. Diagnose it yourself before
opening the solution.
:::

```{r ex2-query}
#| eval: false
sites <- tibble(
  site_id = paste0("SITE_", sprintf("%04d", 1:500)),
  region  = sample(c("North", "South", "East", "West"), 500, replace = TRUE),
  area_ha = runif(500, 10, 500)
)

surveys |>
  left_join(sites, by = "site_id") |>
  filter(region == "North", habitat == "woodland") |>
  group_by(species) |>
  summarise(total = sum(count), .groups = "drop")
```

`r hide("Solution")`

The join precedes the filter, so `left_join` runs on all one million rows before
anything is discarded. `region` lives in `sites`, so it genuinely requires the
join and cannot be filtered earlier. But `habitat` is already in `surveys` and can
be applied first, shrinking the table that enters the join.

```{r ex2-solution}
#| eval: false
bench::mark(
  original = surveys |>
    left_join(sites, by = "site_id") |>
    filter(region == "North", habitat == "woodland") |>
    group_by(species) |>
    summarise(total = sum(count), .groups = "drop"),

  improved = surveys |>
    filter(habitat == "woodland") |>         # reduce surveys before the join
    left_join(sites, by = "site_id") |>
    filter(region == "North") |>             # region needs the join, so filter here
    group_by(species) |>
    summarise(total = sum(count), .groups = "drop"),

  iterations = 10,
  check = FALSE
)[, c("expression", "median", "mem_alloc")]
```

`r unhide()`

---

## 7. Reading the other syntaxes through `show_query()` (≈12 min) {-}

You will write `dplyr`. The lazy back-ends translate it for you, and
`show_query()` lets you read that translation. This is the practical way to meet
`data.table` and SQL: recognise them when they appear, rather than write them from
memory. Full writing references are in Appendices C and D.

```{r show-query}
#| eval: false
# dtplyr translates dplyr to data.table
lazy_dt(as.data.table(surveys)) |>
  filter(habitat == "woodland") |>
  group_by(species, year) |>
  summarise(total = sum(count), .groups = "drop") |>
  show_query()

# dbplyr translates dplyr to SQL (here, DuckDB's dialect)
tbl(con, "surveys_parquet") |>
  filter(habitat == "woodland") |>
  group_by(species, year) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  show_query()
```

The same operation in all three syntaxes:

| `dplyr` | `data.table` | SQL |
|---|---|---|
| `filter(habitat == "woodland")` | `DT[habitat == "woodland"]` | `WHERE habitat = 'woodland'` |
| `select(species, count)` | `DT[, .(species, count)]` | `SELECT species, count` |
| `group_by(a) \|> summarise(n = n())` | `DT[, .(n = .N), by = a]` | `GROUP BY a ... COUNT(*) AS n` |
| `arrange(desc(x))` | `DT[order(-x)]` | `ORDER BY x DESC` |

In `data.table`, the template is `DT[i, j, by]`: `i` filters rows, `j` selects or
computes, `by` groups. In SQL, `WHERE` filters before grouping and `HAVING` filters
after it.

:::{.callout-tip}
**Quick read.** What does each line below do? Say it in words before moving on; you
are practising reading, not writing.

    surveys_dt[year >= 2015 & count > 0, .(total = sum(count)), by = species]

    SELECT species, SUM(count) AS total FROM surveys
    WHERE year >= 2015 AND count > 0 GROUP BY species
:::

:::{.callout-note}
**Instructor.** Show one operation in all three forms, run `show_query()` to confirm
the translation matches, then hand out the reference table and move on. Do not work
through every row; participants return to the table on their own.
:::

---

## 8. Verifying code you did not write line by line (≈10 min) {-}

Four quick checks catch most problems before you run anything on a full dataset.
They apply to any code you are adopting rather than authoring, whatever its source.

**Check 1: the output class.** Confirm the result is the type your downstream code
expects.

```{r check-class}
#| eval: false
class(read_csv("bird_survey.csv", show_col_types = FALSE))
# "tbl_df" "tbl" "data.frame"           a tibble, in memory

class(open_dataset("bird_survey.parquet"))
# "FileSystemDataset" ...               a query description, NOT in memory
```

**Check 2: the translation.** Run `show_query()` before `collect()` on any
`dtplyr` or `dbplyr` pipeline, and check the filter conditions, grouping variables,
and aggregations against your intent.

**Check 3: results agree across engines.** Engines differ on edge cases. The most
common surprise is missing values: `dplyr`'s `sum()` returns `NA` if any input is
`NA`, whereas SQL aggregation functions, DuckDB included, drop `NULL` by default.
For a dataset with no missing counts the two agree, but that coincidence is not a
check. Verify on a small subset.

```{r check-agreement}
#| eval: false
small <- slice_head(surveys, n = 50000)
duckdb_register(con, "small", small)         # an in-memory data frame: register is correct here

reference <- small |>
  filter(habitat == "woodland") |>
  group_by(species, year) |>
  summarise(total = sum(count), .groups = "drop") |>
  arrange(species, year)

candidate <- tbl(con, "small") |>
  filter(habitat == "woodland") |>
  group_by(species, year) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  arrange(species, year)

all.equal(reference$total, candidate$total)
```

:::{.callout-warning}
The `na.rm = TRUE` in the DuckDB version is not cosmetic. It makes the R side match
SQL's default of dropping missing values. Without it, a single `NA` would diverge
the two results. Always verify on a subset before assuming agreement.
:::

**Check 4: performance on your data.** General claims are approximations. Measure
on your file and your machine before committing.

```{r check-performance}
#| eval: false
bench::mark(
  current  = read_csv("bird_survey.csv", show_col_types = FALSE) |>
    filter(habitat == "woodland") |>
    group_by(species, year) |>
    summarise(total = sum(count), .groups = "drop"),

  proposed = tbl(con, "surveys_parquet") |>
    filter(habitat == "woodland") |>
    group_by(species, year) |>
    summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
    collect(),

  iterations = 5,
  check = FALSE
)[, c("expression", "median", "mem_alloc")]

dbDisconnect(con, shutdown = TRUE)
```

If the proposed approach is not faster on your data, the general recommendation
does not apply to your situation. Understand why before you adopt it.

---

## 9. Close: the one capability, and the checklist (≈7 min) {-}

If you take one thing away, take this: choose the format and the evaluation
strategy deliberately. Parquet plus a lazy read with `collect()` lets the engine
read only the columns and row groups your query needs, and brings only the result
into memory. CSV plus an eager read does neither. Everything else in the session,
the optimisation principles, the three syntaxes, the verification checks, exists to
support or to confirm that decision.

### Closing checklist {-}

- [ ] Format chosen on the mechanism: columnar (Parquet) when queries are
      column-selective or filtered; row-oriented (CSV) only when required by an
      external constraint.
- [ ] Reads are **lazy** with `collect()` when the file is large relative to RAM,
      so only the result enters memory.
- [ ] Filters and column selection appear **early**, before joins and aggregations.
- [ ] Partition key matches the most frequently filtered variable, if partitioning.
- [ ] `class()` confirms the output type matches downstream code.
- [ ] `show_query()` checked on any `dtplyr` or `dbplyr` pipeline before `collect()`.
- [ ] Results verified with `all.equal()` on a small subset before scaling.
- [ ] Performance benchmarked on the actual file, not accepted on a general claim.

---

## Appendix A: Key terms {-}

**Column projection**: reading only the requested columns from disk. Works for
columnar formats such as Parquet; not for row-oriented CSV.

**Columnar storage**: a layout in which each column occupies a contiguous block,
independent of other columns. Parquet is columnar.

**`collect()`**: triggers a lazy query and returns a tibble. Before it, a query
description; after it, an in-memory object.

**Common Table Expression (CTE)**: the `WITH name AS (...)` form in SQL. It names an
intermediate query so you can refer to it below, like an inline temporary table.
See Appendix D.

**Copy-on-modify**: R's default behaviour of copying an object when you change it.
`data.table` instead modifies **in place** (the `:=` operator), avoiding copies,
which is part of why it is fast and memory-light on large tables.

**`data.table`**: a high-performance data structure and query engine. Uses several
processor cores at once (multithreading) and in-place modification. Template:
`DT[i, j, by]`.

**`DBI`**: R's standard database interface. `dbConnect()`, `dbExecute()`, and
`dbGetQuery()` are DBI functions; DuckDB implements them.

**`dbplyr`**: translates `dplyr` verbs to SQL for DBI-compatible databases,
DuckDB included. `show_query()` exposes the translation.

**`dtplyr`**: translates `dplyr` verbs to `data.table`. Requires `lazy_dt()`.
`show_query()` exposes the translation.

**DuckDB**: an in-process analytical database engine. No server required; it runs
inside your R session. Reads CSV and Parquet directly.

**Eager evaluation**: computation happens immediately. `read_csv()` and
`read_parquet()` are eager.

**Hive partitioning**: a directory-naming convention, `year=2020/`, that encodes a
column's value in the path. Query engines read the value from the directory name.

**I/O**: input/output, the work of reading from and writing to disk. Usually much
slower than in-memory computation.

**In-process engine**: a database engine that runs inside your program, with no
separate server to install or connect to. Contrast with PostgreSQL or MySQL.

**Lazy evaluation**: computation deferred until `collect()`. `open_dataset()` and
`tbl()` are lazy.

**Multithreading**: using several of the processor's cores at once. `fread()` and
`data.table` are multithreaded; `read_csv()` is single-threaded.

**Partition pruning**: skipping partition directories that cannot satisfy a filter
on the partition key. Activates only on the partition key.

**Peak memory allocation**: the maximum RAM used at any point in an operation. Eager
reads hold the whole file; lazy reads hold only the result. Note that
`bench::mark()`'s `mem_alloc` counts R-heap allocation only.

**Row group**: a horizontal batch of rows within a Parquet file. Per-group min/max
statistics enable row-group skipping during filtered reads.

**Row-oriented storage**: a layout in which all columns for a row are stored
together. CSV is row-oriented; column selection does not reduce its disk I/O.

---

## Appendix B: Tool syntax reference {-}

```{r appendix-readr}
#| eval: false
# readr
read_csv("f.csv",
         col_types      = cols(year = col_integer(), count = col_integer()),
         col_select     = c(species, habitat, year, count),
         n_max          = 10000,
         show_col_types = FALSE)
```

```{r appendix-fread}
#| eval: false
# data.table::fread
fread("f.csv",
      select       = c("species", "habitat", "year", "count"),
      nrows        = 10000,
      showProgress = FALSE)
```

```{r appendix-arrow}
#| eval: false
# arrow
read_parquet("f.parquet")
read_parquet("f.parquet", col_select = c(species, count))

open_dataset("f.parquet") |>
  filter(...) |>
  collect()

open_dataset("dir/") |>            # partitioned dataset
  filter(year == 2020) |>          # partition pruning activates
  collect()

write_parquet(data, "f.parquet", compression = "zstd")
write_dataset(data, path = "dir/", partitioning = "year")
```

```{r appendix-duckdb}
#| eval: false
# duckdb
con <- dbConnect(duckdb())

# Files: expose them as views, then query by name.
dbExecute(con, "CREATE VIEW t_csv     AS SELECT * FROM read_csv_auto('file.csv')")
dbExecute(con, "CREATE VIEW t_parquet AS SELECT * FROM read_parquet('file.parquet')")

# In-memory data frames: register directly (no file involved).
duckdb_register(con, "t_mem", in_memory_tibble)

tbl(con, "t_parquet") |>
  filter(...) |>
  show_query() |>    # inspect the SQL
  collect()          # execute -> tibble

dbGetQuery(con, "SELECT ... FROM t_parquet WHERE ...")

# Partitioned Parquet directly in SQL
dbGetQuery(con, "
  SELECT * FROM read_parquet('dir/*/*.parquet', hive_partitioning = true)
  WHERE year = 2020
")

dbDisconnect(con, shutdown = TRUE)
```

:::{.callout-warning}
`duckdb_register()` registers an in-memory R data frame as a view; its third
argument is a data frame, not a file path. To read files, use a view over
`read_csv_auto()` or `read_parquet()`, as above.
:::

---

## Appendix C: writing `data.table` {-}

```
DT[i, j, by]
    i  = filter rows      (like filter())
    j  = select / compute (like select(), mutate(), summarise())
    by = group            (like group_by())
```

Special symbols: `.N` is `n()`, `.()` is shorthand for a list of columns, `:=`
assigns in place, `uniqueN()` is `n_distinct()`, `shift()` is `lag()`/`lead()`.

```{r appendix-dt-examples}
#| eval: false
surveys_dt <- fread("bird_survey.csv", showProgress = FALSE)

# group and summarise
surveys_dt[, .(total = sum(count), n = .N), by = .(species, habitat)]

# filter, then summarise, then sort
surveys_dt[year >= 2015 & count > 0,
           .(total = sum(count), n_sites = uniqueN(site_id)),
           by = .(species, habitat)][order(-total)]

# lag within group (note: shift() respects existing row order)
surveys_dt[order(year),
           prev_count := shift(count, type = "lag"),
           by = .(species, site_id)]
```

Verify any hand-written `data.table` against the `dtplyr` translation of the
equivalent `dplyr`:

```{r appendix-dt-verify}
#| eval: false
lazy_dt(surveys_dt) |>
  filter(year >= 2015, count > 0) |>
  group_by(species, habitat) |>
  summarise(total = sum(count), n_sites = n_distinct(site_id), .groups = "drop") |>
  arrange(desc(total)) |>
  show_query()
```

---

## Appendix D: SQL window functions and CTEs {-}

A **Common Table Expression (CTE)** names an intermediate result so the query below
can refer to it. It makes multi-step SQL readable, much as a named intermediate
object does in an R pipeline.

```sql
WITH step_one AS ( SELECT ... ),
     step_two AS ( SELECT ... FROM step_one )
SELECT * FROM step_two
```

**Window functions** compute across a set of related rows without collapsing them.
They are essential for time series: lags, rankings, and running totals.

```sql
function() OVER (
  PARTITION BY grouping_column    -- like group_by()
  ORDER BY ordering_column        -- sequence within the group
)
```

```{r appendix-window}
#| eval: false
con <- dbConnect(duckdb())
dbExecute(con, "CREATE VIEW surveys AS
                SELECT * FROM read_parquet('bird_survey.parquet')")

# LAG / LEAD: previous and next value within a group
dbGetQuery(con, "
  SELECT species, site_id, year, count,
    LAG(count)  OVER (PARTITION BY species, site_id ORDER BY year) AS prev_count,
    LEAD(count) OVER (PARTITION BY species, site_id ORDER BY year) AS next_count
  FROM surveys
  ORDER BY species, site_id, year
")

# Top-n per group via a CTE plus RANK()
dbGetQuery(con, "
  WITH annual_totals AS (
    SELECT species, year, SUM(count) AS total
    FROM surveys
    GROUP BY species, year
  )
  SELECT species, year, total,
    RANK() OVER (PARTITION BY year ORDER BY total DESC) AS rnk
  FROM annual_totals
  QUALIFY rnk <= 3
  ORDER BY year, rnk
")

dbDisconnect(con, shutdown = TRUE)
```

`RANK()` gives ties the same rank and then skips positions (1, 2, 2, 4);
`ROW_NUMBER()` gives unique positions regardless of ties (1, 2, 3, 4). The `dplyr`
equivalent of `RANK()` is `min_rank()`, not `rank()`: base `rank()` averages ties
and returns fractional ranks, which disagrees with SQL on tied values.

---

## References {-}

* Trisovic, A., Lau, M. K., Pasquier, T., & Crosas, M. (2022). A large-scale study
  on research code quality and execution. *Scientific Data*, 9, 60.
* Kellner, K. F., Doser, J. W., & Belant, J. L. (2025). Functional R code is rare
  in species distribution and abundance papers. *Ecology*, 106, e4475.
* Apache Arrow R documentation: <https://arrow.apache.org/docs/r/>
* DuckDB R API: <https://duckdb.org/docs/stable/clients/r>
* `dtplyr` documentation: <https://dtplyr.tidyverse.org>
* `dbplyr` SQL translation: <https://dbplyr.tidyverse.org/articles/sql-translation.html>
* `bench` package: <https://bench.r-lib.org>
