# Dates {#sec-dates}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_clean_names <- readRDS("files/penguins_clean_names.RDS")
```

:::{.callout-important collapse = "true"}

## Penguin clean names dataset

```{r, eval = FALSE}
penguins_clean_names <- readRDS(url("https://github.com/UEABIO/5023B/raw/refs/heads/2026/files/penguins.RDS"))

```

:::


Working with dates can be tricky, treating date as strictly numeric is problematic, it won't account for number of days in months or number of months in a year. 

Additionally there's a lot of different ways to write the same date:

* 13-10-2019

* 10-13-2019

* 13-10-19

* 13th Oct 2019

* 2019-10-13

This variability makes it difficult to tell our software how to read the information, luckily we can use the functions in the `lubridate` package. 

::: {.panel-tabset group="language"}

## R

```{r}
library(lubridate)
# library(tidyverse)
```

## Python

```{python}
#| eval: false
import pandas as pd
import numpy as np
from plotnine import *
import statsmodels.formula.api as smf
```

:::

:::{.callout-warning}
If you get a warning that some dates could not be parsed, then you might find the date has been inconsistently entered into the dataset.

Pay attention to warning and error messages
:::

Depending on how we interpret the date ordering in a file, we can use `ymd()`, `ydm()`, `mdy()`, `dmy()` 

## Reformat

We can use functions in the `lubridate` package @R-lubridate to reformat dates written in different ways to *YYYY-MM-DD* format.

::: {.panel-tabset group="language"}

### R

```{r}
date("2017-10-11T14:02:00")
```

### Python

```{python}
#| eval: false
pd.to_datetime("2017-10-11T14:02:00").normalize()
```

:::

::: {.panel-tabset group="language"}

### R

```{r}
dmy("11 October 2020")
```

### Python

```{python}
#| eval: false
pd.to_datetime("11 October 2020", dayfirst=True)
```

:::

::: {.panel-tabset group="language"}

### R

```{r}
mdy("10/11/2020")
```

### Python

```{python}
#| eval: false
pd.to_datetime("10/11/2020")
```

:::

Sometimes dates are presented in formats that do not translate cleanly into R. For example, the following date format isn’t converted correctly when we try to convert it to a date.

::: {.panel-tabset group="language"}

### R

```{r, eval = F}
df <- tibble(
  date = c("X2020.01.22",
           "X2020.01.22",
           "X2020.01.22",
           "X2020.01.22")
)

df |> 
  mutate(
    date = as_date(date)
  )

```

### Python

```{python}
#| eval: false
df = pd.DataFrame({
    "date": ["X2020.01.22", "X2020.01.22", "X2020.01.22", "X2020.01.22"]
})

# TRANSLATION-NOTE: OQ-012
df.assign(date=lambda d: pd.to_datetime(d["date"], errors="coerce"))
```

:::

```
Warning: There was 1 warning in `mutate()`.
ℹ In argument: `date = as_date(date)`.
Caused by warning:
! All formats failed to parse. No formats found.

# A tibble: 4 × 1
  date  
  <date>
1 NA    
2 NA    
3 NA    
4 NA   

```

We can use `%` to be more explicit about what information is in each part of our date column, specifying where the 4-digit year (`%Y`), 2-digit month (`%m`) and 2 digit day (`%d`) are within each string.

::: {.panel-tabset group="language"}

### R

```{r, eval = F}
df |> 
  mutate(
    date = as_date(date, format = "X%Y.%m.%d")
  )

```

### Python

```{python}
#| eval: false
df.assign(date=lambda d: pd.to_datetime(d["date"], format="X%Y.%m.%d"))
```

:::

## Extract

Sometimes we might need to extract certain elements of a longer date-time value for summarising, filtering, or plotting data.

::: {.panel-tabset group="language"}

### R

```{r}
year("2017-11-28T14:02:00")
```

### Python

```{python}
#| eval: false
pd.to_datetime("2017-11-28T14:02:00").year
```

:::

::: {.panel-tabset group="language"}

### R

```{r}
month("2017-11-28T14:02:00")
```

### Python

```{python}
#| eval: false
pd.to_datetime("2017-11-28T14:02:00").month
```

:::

::: {.panel-tabset group="language"}

### R

```{r}
week("2017-11-28T14:02:00")
```

### Python

```{python}
#| eval: false
pd.to_datetime("2017-11-28T14:02:00").isocalendar().week
```

:::

::: {.panel-tabset group="language"}

### R

```{r}
day("2017-11-28T14:02:00")
```

### Python

```{python}
#| eval: false
pd.to_datetime("2017-11-28T14:02:00").day
```

:::

## Excel date formats

Excel stores dates internally as serial numbers — counting the number of days since a specific “origin” date.

- On Windows, the default origin is 1899-12-30

This means that sometimes the date is imported into R as the numeric (number of days since origin). 

The janitor package has a handy function `janitor::excel_numeric_to_date()` to deal with this. 

::: {.panel-tabset group="language"}

### R

```{r}
library(janitor)

excel_numeric_to_date(42370)
```

### Python

```{python}
#| eval: false
pd.to_datetime(42370, unit="D", origin="1899-12-30")
```

:::

:::{.task}
::::{.task-header}
Your turn
::::
::::{.task-container}

* [ ] **Question** What is the appropriate function from the above to use on the `date_egg` variable in the penguins dataset?


`r longmcq(c("ymd()", "ydm()", "mdy()", answer="dmy()"))`


`r hide("Solution")`


:::::{.panel-tabset group="language"}

### R

```{r, eval = T, warning = F, message = F}

penguins_clean_names <- penguins_clean_names |>
  mutate(date_egg = lubridate::dmy(date_egg))

```

### Python

```{python}
#| eval: false
penguins_clean_names = penguins_clean_names.assign(date_egg=lambda d: pd.to_datetime(d["date_egg"], dayfirst=True))
```

:::::

`r unhide()`


Here we use the `mutate` function from `dplyr` to create a *new variable* called `date_egg_proper` based on the output of converting the characters in `date_egg` to date format. The original variable is left intact, if we had specified the "new" variable was also called `date_egg` then it would have overwritten the original variable. 

::::
:::

Once we have established our date data, we are able to perform calculations or extract information. Such as the date range across which our data was collected.  

### Calculations with dates

::: {.panel-tabset group="language"}

#### R

```{r, eval = F}
penguins_clean_names |> 
  summarise(min_date=min(date_egg),
            max_date=max(date_egg))
```

#### Python

```{python}
#| eval: false
penguins_clean_names.agg(min_date=("date_egg", "min"), max_date=("date_egg", "max"))
```

:::

We can also extract and make new columns from our date column - such as a simple column of the year when each observation was made:

::: {.panel-tabset group="language"}

#### R

```{r, eval = F}
penguins_clean_names <- penguins_clean_names |> 
  mutate(year = lubridate::year(date_egg))

```

#### Python

```{python}
#| eval: false
penguins_clean_names = penguins_clean_names.assign(year=lambda d: d["date_egg"].dt.year)
```

:::

### Filter dates

We can filter datasets to include or exclude data from certain dates or date ranges.

::: {.panel-tabset group="language"}

#### R

```{r, eval = FALSE}
# return records after 2008
penguins_clean_names |>
  filter(date_egg >= ymd("2008-01-01"))
```

#### Python

```{python}
#| eval: false
# return records after 2008
penguins_clean_names.loc[lambda d: d["date_egg"] >= pd.to_datetime("2008-01-01")]
```

:::

