# Duplicates {#sec-duplicate}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_clean_names <- readRDS("files/penguins_clean_names.RDS")
```

It is very easy when inputting data to make mistakes, copy something in twice for example, or if someone did a lot of copy-pasting to assemble a spreadsheet (yikes!). We can check this pretty quickly

:::{.callout-important collapse = "true"}

## Penguin clean names dataset

```{r, eval = FALSE}
penguins_clean_names <- readRDS(url("https://github.com/UEABIO/5023B/raw/refs/heads/2026/files/penguins.RDS"))

```

:::


## Duplicated rows


::: {.panel-tabset group="language"}

## dplyr

```{r, eval = F}
library(tidyverse)
# check for whole duplicate 
# rows in the data
penguins_clean_names |> 
  filter(duplicated(across(everything())))
  sum() 

```

```
[1] 0
```


## janitor

```{r}
library(janitor)

penguins_clean_names |> 
  get_dupes()

```

## Python

```{python}
#| eval: false
import pandas as pd
import numpy as np
from plotnine import *
import statsmodels.formula.api as smf

# TRANSLATION-NOTE: OQ-004
# check for whole duplicate rows in the data
penguins_clean_names.duplicated().sum()
```

:::

Great! In our dataset we have no examples of duplicated rows of data

### Working with duplications

As our dataset is duplication free, lets quickly add a few dupes to our data: 

::: {.panel-tabset group="language"}

#### R

```{r}
penguins_demo <- penguins_clean_names |> 
  slice(1:50) |> 
  bind_rows(slice(penguins_clean_names, c(1,5,10,15,30)))

```

#### Python

```{python}
#| eval: false
# TRANSLATION-NOTE: OQ-005
penguins_demo = pd.concat([
    penguins_clean_names.iloc[0:50],
    penguins_clean_names.iloc[[0, 4, 9, 14, 29]],
])
```

:::

:::{.task}
::::{.task-header}
Your turn
::::
::::{.task-container}

* [ ] Run the code above to make a dataset with some duplications

* [ ] Re-run a duplication check

::::
:::

If I did have duplications I could remove these with a few commands: 

:::{.panel-tabset group="language"}

## duplicated

```{r, eval = F}

# Keep only unduplicated data with !
penguins_demo |> 
  filter(!duplicated(across(everything())))

```

## distinct

```{r, eval = F}


penguins_demo |> 
  distinct()

```

## Python

```{python}
#| eval: false
penguins_demo.drop_duplicates()
```

:::



### Counting unique entries

Using the `n_distinct()` function from dplyr, you can count the number of distinct values in an R data frame using one of the following methods.


::: {.panel-tabset group="language"}

#### R

```{r}
penguins_clean_names |> 
  summarise(
  n = n(),
  n_distinct(individual_id)
  )

```

#### Python

```{python}
#| eval: false
# TRANSLATION-NOTE: OQ-006
penguins_clean_names.agg(n=("individual_id", "size"), n_distinct=("individual_id", "nunique"))
```

:::




