# Model diagnostics {#sec-assumption}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
library(skimr)
library(ggridges)
library(lmtest)
library(patchwork)
library(performance)
library(parameters)
library(sandwich)
library(car)
library(openintro)
source("R/booktem_setup.R")
source("R/my_setup.R")

```




```{r, eval=F, echo=T}

library(tidyverse)
library(here)
library(performance)
library(parameters)
library(sandwich)
library(car)
library(openintro)

```


:::{.callout-warning}

Set this up as a new R project, with organised folders. 

:::

## Model assumptions


Linear models are a cornerstone of statistical analysis, widely used across scientific disciplines to explore relationships between variables. Their popularity stems from their simplicity, interpretability, and versatility. However, the reliability of linear models depends on meeting several critical assumptions: 

- Normality of residuals

- Homogeneity of variance

- Linearity of relationships

- Absence of influential outliers 

- Low multicollinearity among predictors.

When these assumptions are violated, the model’s conclusions can become misleading, leading to erroneous inferences and predictions. Diagnosing and addressing these violations is a crucial step in any linear modeling workflow. This chapter explores how to evaluate linear models, identify assumption violations, and apply appropriate corrective measures.


## Missing variables

When learning statistics, it’s important to remember that the first step in diagnosing poor model residuals is to ensure that your model includes the correct predictors and interaction terms. Residuals reflect the difference between the observed and predicted values, and large or patterned residuals often indicate that the model is missing key information.

For example:

If a predictor that strongly affects the outcome is omitted, the model may fail to capture important variation, leading to biased predictions.
If an interaction between two predictors (e.g., temperature and humidity) exists but is not included, the model might not adequately describe the relationship between predictors and the outcome.

### Blood Pressure Data

```{r, include = FALSE}

bp <- read_csv("files/bp.csv")


```

This dataset includes key physiological and demographic variables — **Blood Pressure, Weight, Age, Albumin (a blood protein level), Pulse, and Stress**. These variables are commonly studied in health and biological research to understand how physiological and psychological factors interact.

We are specifically interested in the effects of Weight and Blood albumin levels on blood pressure readings: 

```{r}

bp_model <- lm(BP ~ Weight + Albumin, data =bp)
summary(bp_model)
```

```{r}

performance::check_model(bp_model,
                         detrend = F)

```

**What are the issues with these model residuals?**

```{r, echo = F}
opts_p <- c(
   answer = "non-normality of residuals",
   "Heteroscedasticity of variance",
   answer = "Non-linearity",
   "Outliers"
)
```

`r longmcq(opts_p)`


`r hide("Solution")`

There residuals do not fit the assumption of a normal distribution. In addition there are some non-linear patterns in the residuals that indicate this model may not be the best fit

`r unhide()`


### Improving the blood pressure model

One of the simplest methods to improve a model can be to reevaluate whether important predictor variables have been left out. Can you add other predictors to the model and evaluate whether they improve the fit?

`r hide("Solution")`

```{r}
bp_model2 <- lm(BP ~ Weight + Albumin + Age, data =bp)

performance::check_model(bp_model2,
                         detrend = F)
```


From this we can see that adding Age is an important covariate for our model. If we chose to include Pulse or Stress as well, that would be ok - they don't make the model any worse - but they don't appear to be particularly important. 

From this we can learn that before we attempt any radical data transformations, we should check whether our model would fit the data better with the inclusion of an omitted variable. 

When we do this, not only does our model fit the data better, but our understanding of the importance of our predictor variables changes as well. Previously we would have dismised Albumin levels as unimportant, now we see that when account for Age as a predictor of blood pressure, albumin levels are also a significant predictor. 

```{r}

summary(bp_model2)

```

`r unhide()`


## Data transformations

When building a linear model, we assume that the residuals (the differences between the observed and predicted values) are small, randomly distributed, and have constant variance. If this isn’t true, the model may not be capturing the relationship between the variables correctly.

Transforming the dependent variable (the variable you’re trying to predict) can help fix these issues by changing its scale to better match the assumptions of the model. Here’s how:

- Non-normal residuals: If the dependent variable is heavily skewed (e.g., income or population size), transformations like taking the  log(y) or square root (y) can make the distribution more symmetric.

- Stabilize Variance: If the residuals show increasing or decreasing spread (heteroscedasticity), transformations can help make the variance more constant.

- Linearise Relationships: Some relationships between the dependent and independent variables are non-linear. Transformations can make these relationships more linear, which linear models handle better.

### Example

```{r, eval = FALSE, include = FALSE}

# Set random seed for reproducibility
set.seed(123)

# Number of observations
n <- 100

# Generate a continuous predictor: temperature (degrees Celsius)
temperature <- runif(n, min = 20, max = 40)

# Generate a categorical predictor: nutrient type (A, B, C)
nutrient_type <- sample(c("A", "B", "C"), n, replace = TRUE)

# Simulate bacterial growth rate with an exponential relationship to temperature
# and additive effects of nutrient type
growth_rate <- rnorm(n, mean = 0.9, sd = .3) * exp(0.2 * temperature) 
 
  

# Combine into a data frame
data <- data.frame(
  Temperature = temperature,
  Growth_Rate = growth_rate
)

# View the first few rows of the dataset
head(data)

# Optionally save or visualize
# write.csv(data, "bacterial_growth.csv", row.names = FALSE)




```


```{r, include = FALSE}

bacteria_growth_data <- read_csv("files/bacterial_growth.csv")

```


```{r}

# Model and diagnostics
bacteria_model <- lm(Growth_Rate ~ Temperature, data = bacteria_growth_data)

check_model(bacteria_model,
            detrend = FALSE)

```

** What are the issues with this model?**

`r hide("Solution")`

If we look at the residuals we see there are obvious issues with linearity, and heteroscedasticity and normality of the residuals - all of these criteria *might* be improved with the right data transformation

```{r}

check_normality(bacteria_model)

check_heteroscedasticity(bacteria_model)

```

`r unhide()`

#### Boxcox

Box-Cox transformations help identify the best transformation for the dependent variable (y) in a linear model to improve fit, stabilize variance, and address non-normality. The method tests a family of power transformations:

 
How It Works:
The Box-Cox method estimates the optimal value of 𝜆 (the power parameter) that makes the residuals closest to normality and stabilizes variance.

Common transformations include:

- 𝜆= 1: No transformation (original scale).

- λ=0: Log transformation.

- λ=0.5: Square root transformation.


:::{callout-note}

Why Not Use 𝜆 Directly?
The optimal 𝜆is often not a simple value like 0, 0.5, or 1. 
Using a fractional 𝜆(e.g., 0.27) creates a less interpretable transformation.
Instead of applying the exact 𝜆 it’s better to choose a nearby, familiar transformation (e.g., log or square root) for ease of interpretation while still improving the model.

:::

```{r}

car::boxCox(bacteria_model)

```




```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
library(kableExtra)

`lambda value` <- c(0, 0.5, 1, 2 )

`transformation` <- c("log(Y)", "sqrt(Y)", "Y", "Y^1")

box <- tibble(`lambda value`, `transformation`)

box %>% 
   kbl(caption = "Common Box-Cox Transformations", 
    booktabs = T) %>% 
   kable_styling(full_width = FALSE, font_size=16)

```


**What is the optimal transformation for our dependent variable according to the Box-Cox transformation**

```{r, echo =F}
opts_p <- c(
    "Identitity/No transformation",
   "Square root",
   answer = "Logarithm",
   "Inverse"
)
```

`r longmcq(opts_p)`


`r hide("Solution")`


```{r}


# Model and diagnostics
log_bacteria_model <- lm(log(Growth_Rate) ~ Temperature, data = bacteria_growth_data)

check_model(log_bacteria_model,
            detrend = FALSE)

check_normality(log_bacteria_model)

check_heteroscedasticity(log_bacteria_model)

```

`r unhide()`


### Potential drawbacks

Transforming the dependent variable can help fix issues like skewed data or uneven residual patterns, but it can also create new challenges:

- Interpretability:

After a transformation, the relationship between predictors and the dependent variable becomes harder to interpret. For example, if you log-transform the dependent variable, the model predicts the logarithm of y, not y itself. This makes it less intuitive to explain results on the original scale.

- Fixing One Issue, Causing Another:

While a transformation might solve one problem (e.g., skewed data), it could create new issues. For instance, a log transformation can stabilize variance but may introduce difficulties if the data contains zero or negative values.

In short, while transformations are useful, they require careful consideration to avoid trading one problem for another or making the results harder to interpret.


## Heteroscedasticity

Heteroscedasticity occurs when the residuals (errors) of a linear model have non-constant variance across the range of the predictors. Even if the data appears normal and linear, heteroscedasticity can still be a problem.

Imagine a situation where:

- Residuals are larger (more spread out) for higher values of a predictor, or smaller for lower values.

- Although the relationship between the predictors and the response variable is linear, the variance of the errors changes systematically.

#### Why Does This Matter?

Misleading Inference: Heteroscedasticity doesn’t bias the coefficients of the model, but it affects the calculation of standard errors, leading to incorrect p-values and confidence intervals.

```{r, include = FALSE}

# Set random seed for reproducibility
set.seed(123)

# Number of observations
n <- 100

# Generate a predictor variable
sunlight<- runif(n, min = 1, max = 10)

water <- runif(n, min=1, max = 20)

# Generate heteroscedastic errors: larger variance for larger values of the predictor
errors <- rnorm(n, mean = 0, sd = sunlight*4)

# Generate the response variable with a linear relationship to the predictor
plant_growth <- 30 + 3 * sunlight + .2*water + errors

# Combine into a data frame
plant_data <- data.frame(
  sunlight = sunlight,
  water = water,
  plant_growth = plant_growth
)

# View the first few rows of the dataset
head(data)

# Optionally save or visualize
# write.csv(plant_data, "files/sunlight_water_data.csv", row.names = FALSE)
```


```{r, include = FALSE}

plant_data <- read_csv("files/sunlight_water_data.csv")

```


```{r}

# Model and diagnostics
plant_model <- lm(plant_growth ~ sunlight + water, data = plant_data)

check_model(plant_model,
            detrend = FALSE)

check_heteroscedasticity(plant_model)
check_normality(plant_model)

```


**What are the issues with these model residuals?**

```{r, echo =F}
opts_p <- c(
   "non-normality of residuals",
   answer = "Heteroscedasticity of variance",
   "Non-linearity",
   "Outliers"
)
```

`r longmcq(opts_p)`


#### Transformation

Unlike the previous example we have heteroscedasticity of variance, without non-linearity or non-normal residuals. So what happens when we attempt a dependent variable transformation?

First let's see what a box-cox transformation suggests?

```{r, eval = FALSE}

car::boxCox(plant_model)

```

**Why does the Box-Cox transformation throw an error when the dependent variable contains non-positive values (zero or negative)?**

```{r, echo = F}
opts_p <- c(answer = "The transformation involves logarithms, which are undefined for non-positive values.", "The dependent variable must be normally distributed before applying Box-Cox.", "Box-Cox transformations cannot handle decimals or fractions in the dependent variable.", "Non-positive values violate the assumption of homoscedasticity required for Box-Cox."
)
```

`r longmcq(opts_p)`

`r hide("Solution")`

If we add a **constant** to the dependent variable (enough to make all values positive), the relationship between dependent and independent variables is preserved but the all values are now non-negative.

```{r}

plant_model <- lm(plant_growth+1 ~ sunlight + water, data = plant_data)

car::boxCox(plant_model)

```

This indicates a square root transformation could be applied to the model


```{r}

sqrt_plant_model <- lm(sqrt(plant_growth+1) ~ sunlight + water, data = plant_data)

check_model(sqrt_plant_model,
            detrend = FALSE)

check_heteroscedasticity(sqrt_plant_model)
check_normality(sqrt_plant_model)

```

Despite the box-cox predictions, we should be able to see that transforming the dependent variable has actually made this worse. 

Not totally unexpected, a data transformation changes the relationship between the predictor variables and the mean/ regression line as well as affecting variance. 


`r unhide()`


### Robust errors

When the only issue with a model is heteroscedasticity of variance, the relationship between predictors and the dependent variable can be good, but our standard errors, confidence intervals and p-values will be unreliable. 

In a linear model, **robust errors** are adjustments to the standard errors of the model's coefficients that make them more reliable when certain assumptions, like constant variance of residuals (homoscedasticity), are violated (giving heteroscedasticity).

Imagine you’re analysing data where some observations have more variability than others. Ordinary linear models assume all residuals have the same variance, so they might underestimate (or overestimate) the uncertainty of your coefficients. Robust errors correct this by accounting for the uneven variance, giving you more reliable p-values and confidence intervals.

In R we can use the package @R-sandwich to help provide robust errors: 

```{r}

coeftest(plant_model, 
         vcov = vcovHC(plant_model, type = "HC1"))

```

:::{callout-warning}

The emmeans package doesn't support the sandwich model outputs directy, but we can supply the same adjustments when running the emmeans function

```{r, eval = FALSE}

# Get robust covariance matrix
robust_vcov <- vcovHC(plant_model, type = "HC1")

# Use emmeans with robust covariance
# specify lists to get predictions for continuous variables
emmeans::emmeans(plant_model, specs = ~ sunlight + water, 
                 at = list(sunlight = c(1:10),
                           water = 1:20),
                 vcov. = robust_vcov)

```

:::


## Bootstrapping

Bootstrapping is a statistical technique that uses resampling to estimate the variability of a model's predictions or parameters, such as coefficients and standard errors. Unlike traditional methods that rely on strict assumptions about the data (e.g., normality or constant variance), bootstrapping works by repeatedly sampling from the observed data or residuals to create many simulated datasets we create many versions of what the data "could have looked like."

This process reflects the natural variability in the data, giving us a realistic measure of uncertainty.

By fitting the model to these resampled datasets, we can calculate how often our estimates and standard errors change - if they are generally constant - we can have confidence they are useful even when they may violate some of the assumptions of the standard linear model. 

### Things bootstrapping can help with: 

1. Non-Normality of Residuals

    - Problem: Linear models assume that residuals (errors) are normally distributed. If this assumption is violated (e.g., due to skewed data or outliers), the standard errors and confidence intervals based on the normality assumption may be invalid.

    - How Bootstrapping Helps: Bootstrapping does not rely on the residuals being normally distributed. Instead, it estimates variability directly by resampling from the data, providing robust confidence intervals and standard errors even when residuals are non-normal.

2. Heteroscedasticity (Non-Constant Variance of Residuals)

    - Problem: Linear models assume that residuals have constant variance (homoscedasticity). If the variance of residuals changes across levels of a predictor, standard errors can be underestimated or overestimated.
    
    - How Bootstrapping Helps: Bootstrapping accounts for heteroscedasticity by resampling residuals with their original variability, providing standard errors and confidence intervals that reflect the true spread of the data.

3. Small Sample Sizes

    - Problem: Small datasets make it hard to rely on traditional statistical methods, as they often assume large sample sizes for accurate standard errors and confidence intervals.

    - How Bootstrapping Helps: By repeatedly resampling the available data, bootstrapping effectively "amplifies" the sample size, allowing for more stable estimates of variability even with small datasets.

4. Outliers or Influential Points

    - Problem: Outliers or influential data points can skew the model fit, affecting estimates and residual patterns.

    - How Bootstrapping Helps: Bootstrapping resamples the data, ensuring that results are not overly dependent on specific observations. It gives a more robust estimate of variability that accounts for the influence of outliers.

### Limitations of Bootstrapping

While bootstrapping is powerful, it does not address all issues, it cannot help with: 

- Multicollinearity

- Missing Key Predictors: If an important predictor is omitted, bootstrapping will not fix the resulting bias in the model.

- Non-linearity: Bootstrapping does not correct for non-linear relationships; re-specifying the model is necessary.

:::{.callout-tip}

Bootstrapping can be very powerful, but it doesn't fix everything (see above). 

In addition bootstrapping does not teach us about the real patterns of variance or distribution of residuals in the data. Instead, it works by resampling the observed residuals (or data) and assumes that these resampled patterns are representative of the true variability.

Finally it is computationally intensive as the data is resampled and modelled many times

:::

We can fit bootstrapped parameters using the @R-parameters package: 

- We usually set the iterations argument to a minimum of 1000 iterations, it may need more. You have to bootstrap until the estimates are *stable* (appear not to change as iterations increase)

- If the estimates change only negligibly (e.g., within 1–2%), you have likely run enough iterations.


```{r}

parameters::bootstrap_parameters(plant_model, iterations = 1000)

# parameters::bootstrap_parameters(plant_model, iterations = 3000)
```

Because bootstrapping is inherently random, you can get slightly different results each time it is run, for that reason we often use the `set.seed()` function - this function is used to create reproducible results when writing code that involves random number generation.

```{r, eval = FALSE}
set.seed(123)

times <- c(500,1000,3000,5000, 10000)

# This code iterates the bootstrapping over increasing iterations
times |> 
  map(~ bootstrap_parameters(model = plant_model, iterations = .x))

```


```
[[1]]
# Fixed Effects

Parameter   | Coefficient |         95% CI |      p
---------------------------------------------------
(Intercept) |       47.28 | [33.13, 59.40] | < .001
sunlight    |        0.02 | [-1.68,  1.80] | 0.980 
water       |       -0.02 | [-0.81,  0.90] | 0.980 

[[2]]
# Fixed Effects

Parameter   | Coefficient |         95% CI |      p
---------------------------------------------------
(Intercept) |       47.87 | [33.49, 62.12] | < .001
sunlight    |   -7.41e-04 | [-1.74,  1.66] | > .999
water       |       -0.04 | [-0.87,  0.95] | 0.938 

[[3]]
# Fixed Effects

Parameter   | Coefficient |         95% CI |      p
---------------------------------------------------
(Intercept) |       47.71 | [32.81, 61.19] | < .001
sunlight    |       -0.02 | [-1.74,  1.67] | 0.979 
water       |   -3.32e-03 | [-0.86,  0.91] | 0.994 

[[4]]
# Fixed Effects

Parameter   | Coefficient |         95% CI |      p
---------------------------------------------------
(Intercept) |       47.70 | [33.16, 61.24] | < .001
sunlight    |   -6.85e-03 | [-1.67,  1.71] | 0.995 
water       |   -1.45e-03 | [-0.86,  0.87] | 0.998 

[[5]]
# Fixed Effects

Parameter   | Coefficient |         95% CI |      p
---------------------------------------------------
(Intercept) |       47.63 | [33.88, 61.09] | < .001
sunlight    |       -0.01 | [-1.70,  1.69] | 0.988 
water       |   -7.96e-03 | [-0.86,  0.86] | 0.986 

```


## Non-linearity

In linear models, we assume that the relationship between the predictors (independent variables) and the response (dependent variable) is linear—that is, a change in a predictor is associated with a proportional change in the response. However, real-world data often deviate from this assumption, and relationships can be non-linear. For example, the effect of increasing temperature on plant growth might plateau after an optimal range, or the impact of exercise on health might diminish after a certain intensity.

Let's revisit the janka data from our introductory regression chapter

```{r, eval=T, echo = FALSE}
library(SemiPar)
data(janka)

```

We originally fitted a simple regression model of wood density as a predictor of timber hardness: 

```{r}

janka_ls1 <- lm(hardness ~ dens, data = janka) 

summary(janka_ls1)

```

If we inspect this model we can see there are issues with non-linearity, heteroscedasticity and non-normality: 

```{r}

performance::check_model(janka_ls1)

```


### Transformation

**Which model does a Boxcox transformation suggest is most appropriate?**

`r hide("Solution")`

```{r}

car::boxCox(janka_ls1)

```

`r unhide()`

```{r, echo = F}
opts_p <- c(
   "Identity",
   "square root",
   answer = "log",
   "inverse"
)
```

`r longmcq(opts_p)`

We can fit both a sqrt and log transformed model and compare them: 

```{r}

janka_sqrt <- lm(sqrt(hardness) ~ dens, data = janka)

janka_log <- lm(log(hardness) ~ dens, data = janka)

```


```{r}

performance::check_model(janka_sqrt)

```
This model does not significantly improve the normality or heterogeneity of the residuals, but does improve linearity


```{r}

performance::check_model(janka_log)

```

This model does improve normality and heterogeneity, but introduces an overcorrection on linearity  

### Polynomials

Polynomials are mathematical expressions that allow us to model non-linear relationships between predictors and the response variable in a linear model. By including polynomial terms (e.g., x^2, x^3), we can create more flexible models to capture curvature in the data.

#### What Does `poly(x, 2)` Do?

- `poly(x, 2)` creates a **second-degree polynomial** (quadratic).
- It models the relationship as:
  \[
  y = \beta_0 + \beta_1 x + \beta_2 x^2 + \epsilon
  \]
- This adds curvature to the model, allowing it to capture a U-shape or an inverted U-shape.

##### Example:

If the data looks like a parabola (e.g., plant growth increasing with temperature up to a point, then declining), a second-degree polynomial can model this.


#### What Does `poly(x, 3)` Do?
- `poly(x, 3)` creates a **third-degree polynomial** (cubic).

- It models the relationship as:
  \[
  y = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \epsilon
  \]
  
- This allows for more flexibility in the curve, capturing an S-shaped pattern or two bends in the data.

##### Example:

If the relationship involves multiple changes in direction (e.g., population growth speeding up, then slowing down, and finally stabilizing), a cubic polynomial can model this.

#### Janka data

In our example the slight curvature in the model could be addressed with a quadratic term: 

```{r}

janka_poly<- lm(hardness ~ poly(dens,2), data = janka)


```

But we can see this does not address the issues of heteroscedasticity or non-normality of residuals.

```{r}

performance::check_model(janka_poly)

```

We know that the log transformed dependent variable did address this, but introduced unwarranted curvature - we could address this with a polynomial:

```{r}

janka_polylog<- lm(log(hardness) ~ poly(dens,2), data = janka)


```


```{r}

performance::check_model(janka_polylog)

```

This produces a much better fitting model - but what about interpretation?

```{r}

summary(janka_polylog)

```

We now have an excellent fitting model - but this has come at the cost of interpretation. 

- The coefficients are not directly interpretable

- The relationship with the dependent is on a log scale


#### Polynomial summary

Polynomials can help improve the fit of a model when there is a non-linear relationship. When you do this focus on the overall model fit and visualizations, as the coefficients from `poly()` are not directly tied to raw powers of x. Because of the lack of coefficient interpretation use this with *caution*

- poly(x, 2) (Quadratic): Adds curvature, useful for simple bends like U-shapes.

- poly(x, 3) (Cubic): Adds more flexibility, useful for S-shaped or complex curves.

In our advanced modules we will look at **Generalised Linear Models** which retain the benefits of modelling non-linear relationships with better interpretability


## Multicollinearity

Multicollinearity occurs when predictor variables in a linear model are highly correlated, which can make it hard to isolate the effect of individual predictors.

If the primary aim of our model is **prediction** we do not need to be concerned about multicollinearity as our model will predict new data well

If our aim is to identify **significant predictors** then multicollinearity makes it difficult to produce accurate estimates and standard errors:

However, not all types of multicollinearity need to be a concern.

### Structural Multicollinearity

- **What it is**: This happens when we create new variables (e.g., interaction terms or quadratic terms) that are mathematically related to the original variable naturally introduces multicollinearity

- **Do we need to worry?**: No, structural multicollinearity is expected and doesn’t affect the validity of the model. It can often be reduced by mean-centering or standardising the variables, but this is often unnecessary.

### Predictor Multicollinearity

- **What it is**: This occurs when two or more predictors are strongly correlated in the dataset (e.g., temperature and humidity in an environmental study). This can inflate standard errors and make it difficult to interpret individual effects.

- **Do we need to worry?**: Sometimes. If your goal is prediction, multicollinearity is less of an issue because the model’s overall performance is what matters. However, if you’re interested in understanding the effect of individual predictors, you may need to address it.

Predictor Multicollinearity can be addressed using simple techniques like variable selection, where we selectively remove terms from the model, this can address inflated standard errors, but requires careful thought over removing terms. More advanced technqiues include principal components analysis (PCA), or regularization (e.g., ridge regression). 

### Example

The Mammals dataset includes information about various mammal species, focusing on their physical characteristics and ecoogicial traits. 

```{r}
library(openintro)

data("mammals")

```


We will look at the follwing traits: 

1. **Physiological Variables**:

   - `total_sleep`: The average number of hours a species sleeps per day.
   - `life_span`: The average lifespan of the species in years.
   - `gestation`: The average gestation period (pregnancy duration) in days.
   - `brain_wt`: Brain weight of the species in kilograms.
   - `body_wt`: Body weight of the species in kilograms.

2. **Ecological and Behavioral Variables**:

   - `predation`: A categorical variable indicating the species' risk of predation (e.g., low, medium, high).
   - `exposure`: A categorical variable describing how exposed the species is while sleeping.
   - `danger`: A categorical variable summarizing the overall danger level to the species.


#### Correlations

We can get a good idea of which variables might produce multicollinearity by first seeing which are highly correlated: 

```{r}
mammals |> 
  dplyr::select(total_sleep, life_span, gestation, brain_wt,
               body_wt, predation, exposure, danger) |> 
  GGally::ggpairs()

```

**Which sets of variables have the highest correlations?**

`r hide("Solution")`

- Danger and predation

- Brain weight and Body weight

`r unhide()`

### Fitting a model

I am fitting a model to establish the behavioural and physiological characteristics that predict sleep:

```{r}
mammal_model <- lm(total_sleep ~ life_span + gestation + brain_wt +
               body_wt + predation + exposure + danger, data=mammals)

summary(mammal_model)
```

The model summary suggests that "danger" and "predation" are the most likely predictors of sleep duration

The model residual checks indicate this is an excellent model fit but with high multicollinearity: 

```{r}
performance::check_model(mammal_model)

```

### Correcting for multicollinearity

In this instance we have no structural multicollinearity and must conclude there is genuine multicollinearity between our measured variables. 

There are advanced techniques for dealing with predictor multicollinearity, but the simplest method is to identify the variables causing mulitcollinearity and use your biological expertise to determine which variables can be removed.

In this instance I should remove one of predation/danger and one of brain_wt/body_wt.

I have decided that overall danger and brain_wt are the most important variables so will retain these: 

```{r}

mammal_model2 <- lm(total_sleep ~ life_span + gestation + brain_wt +
                   danger + exposure, data=mammals)

performance::check_model(mammal_model2)

```

I can see that my multicollinearity has gone AND my model is still a good fit.

Importantly if I compare the summaries of my old and new models, I can see that my **adjusted R squared** has only dropped slightly and that gestation period now appears to be a significant predictor of sleep duration - previously this relationship was masked!

```{r}

summary(mammal_model2)
```

When reporting the effects of multicollinearity it is important to:

- Report variables that were removed for reasons of multicollinearity

- Compare the two models for fit

## Outliers

Outliers have been left to last because they should be dealt with last. Outliers can be the result of a poor model fit, non-linearity or non-normal residuals. As such they may be resolved when attempting to fix other issues with a model. 

```{r, include = FALSE}

titanosaurs <- read_csv("files/titanosaurs.csv")

```
Cope’s rule refers to the tendency of animal body sizes to increase in time with increasing age of the lineage. De Souza and Santucci (2014) tested this using a set of data from the dinosaur clade called the Titanosaurs. This is a group of sauropods that included some of the largest animals known in the history of the Earth (e.g. Argentinosaurus) and which were widespread and diverse during the Cretaceous period. Most Titanosaurs are known from only a few bones, and the massive limb bones are particularly common: In this dataset we wish to see whether there is a significant relationship between two sets of bones, so that if we only have access to the Humerus we can predict the size of the Femur and the overall size of the dinsosaur: 

```{r}

titanosaur_model <- lm(Femur ~ Humerus, data = titanosaurs)

summary(titanosaur_model)
```


```{r}

performance::check_model(titanosaur_model,
                         detrend = FALSE)

```
In this model the residual fits are good (perhaps slight non-linearity, but lets leave that for now). But we have one outlier. 

Assumin all other aspects of the model are good, we can check whether this outlier is affecting our model interpretations by removing it and evaluating the model without it

The diagnostic plot indicates that the data is from **row 1**

The easiest way to address this is to use co-ordinates:

```{r}
# removes row 1 [row, column]
titanosaurs2 <- titanosaurs[-c(1), ]

#refit model
titanosaur_model2 <- lm(Femur ~ Humerus, data = titanosaurs2)

summary(titanosaur_model2)

```

Importantly the model estimate **has**changed, but it is still a significant predictor, and there are no further outliers:

```{r}

performance::check_model(titanosaur_model2,
                         detrend = FALSE)

```

If you remove an outlier causing leverage you must report it, and you should compare the models and report whether it has altered any of your coefficient estimates and significant predictors.



## Advanced Data Analysis

Linear models are a foundational tool in data analysis, providing a straightforward way to understand relationships between variables. However, they are just the beginning. As you progress, more advanced techniques will equip you to tackle a wider range of data and research questions:

- Generalized Linear Models (GLMs): Extend linear models to handle different types of response variables, such as counts (Poisson regression) or proportions (logistic regression).

- Mixed Models: Account for hierarchical or grouped data, such as repeated measures or experiments with nested designs, by including both fixed and random effects.

- Machine Learning Algorithms: Focus on predictive accuracy and uncovering complex patterns in data using techniques like decision trees, random forests, and neural networks.

- Classification: Specialized methods, like logistic regression or support vector machines, are used to predict categorical outcomes (e.g., disease presence or absence).

- Principal Component Analysis (PCA): A dimensionality reduction technique that simplifies large datasets by transforming variables into a smaller set of uncorrelated components while retaining most of the variability.

Each of these methods builds on the principles of linear models, expanding your ability to analyse complex data, make predictions, and extract meaningful insights across diverse contexts. Your journey in data analysis has only just begun!
