## Check Data Consistency {#sec-numeric-consistency}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_clean_names <- readRDS("files/penguins_clean_names.RDS")
```

:::{.callout-important collapse = "true"}

## Penguin clean names dataset

```{r, eval = FALSE}
penguins_clean_names <- readRDS(url("https://UEABIO/5023B/raw/refs/heads/2026/files/penguin_clean_names.RDS"))

```

:::

### Why check numeric consistency?

Numeric data can contain errors that aren't immediately obvious. Values might be:

- Biologically impossible (negative body mass, height exceeding known species maximum)

- Data entry errors (decimal point in wrong place: 3500g recorded as 35g)

- Wrong units (measurements in grams recorded in a column meant for kilograms)

- Implausible combinations (male penguins recorded as laying eggs, heavy body mass with short flippers)

Unlike categorical or date errors which often cause R to throw warnings, implausible numeric values will silently propagate through your analysis, potentially leading to incorrect conclusions.

:::{.callout-warning}
Just because a value is an outlier doesn't mean it's wrong. Similarly, a value within the normal range could still be an error if it's inconsistent with other measurements. You need biological knowledge AND cross-variable checks to distinguish errors from legitimate variation.
:::


### Basic range checks

The first step is to examine the range of your numeric variables and ask: *"Are these values biologically possible?"*

```{r}
# Check ranges of all numeric variables at once
penguins_clean_names |> 
  summarise(across(where(is.numeric), 
                   list(min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE))))

```

Let's focus on body mass:

```{r}
# Check body mass range
penguins_clean_names |> 
  summarise(
    min_mass = min(body_mass_g, na.rm = TRUE),
    max_mass = max(body_mass_g, na.rm = TRUE)
  )

```

### Detecting Impossible values

Check for biologically impossible values - things that cannot exist in nature:

```{r}
# Check for negative values (impossible for mass, length measurements)
penguins_clean_names |> 
  filter(if_any(c(body_mass_g, flipper_length_mm, 
                  culmen_length_mm, culmen_depth_mm), 
                ~ . < 0))

```

```{r}
# Check for zero or negative values where zero doesn't make biological sense
penguins_clean_names |> 
  filter(body_mass_g <= 0)

```

:::{.callout-note}
A body mass of exactly 0g is impossible - this likely represents a missing measurement that was coded as 0 instead of NA. This is a common data entry practice but is problematic for analysis because 0 is treated as a real value, not as missing data.
:::

### Species-specific checks

What's plausible for one species may be implausible for another:

```{r}
# Body mass ranges by species
penguins_clean_names |> 
  group_by(species) |> 
  summarise(
    min_mass = min(body_mass_g, na.rm = TRUE),
    max_mass = max(body_mass_g, na.rm = TRUE),
    mean_mass = mean(body_mass_g, na.rm = TRUE)
  )
```

From the scientific literature, we know:

- Gentoo penguins are the largest (typically 4500-6500g)

- Adelie and Chinstrap are similar in size (typically 3000-4500g)

```{r}
# Find Adelie penguins with Gentoo-sized body mass
penguins_clean_names |> 
  filter(species == "Adelie Penguin (Pygoscelis adeliae)", body_mass_g > 4750)
```

:::{.callout-tip}
When you find a species-implausible value, consider:

Is the measurement wrong?
Is the species identification wrong?
Are the units wrong?
Is this a juvenile or unusual individual (still correct)?
:::


### Visual inspection


```{r}
#| label: fig-mass-boxplot
#| fig-cap: "Body mass distribution by species. Visual inspection reveals outliers and unusual patterns."

penguins_clean_names |> 
  ggplot(aes(x = species, y = body_mass_g)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(
    x = "",
    y = "Body mass (g)"
  ) +
  theme_minimal()+
  coord_flip()

```

### Cross-variable checks: Expected correlations

Individual variables might look fine, but relationships between variables can reveal errors. Body size measurements should correlate positively:

```{r}

#| label: fig-mass-flipper
#| fig-cap: "Body mass should generally increase with flipper length within species. Points far from the trend may indicate measurement errors."

penguins_clean_names |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
  ) +
  theme_minimal()

```
```{r}
# Find penguins with large flippers but low body mass
penguins_clean_names |> 
  filter(flipper_length_mm > 210, body_mass_g < 3500) |> 
  select(species, sex, flipper_length_mm, body_mass_g, island)

```


```{r}
# Find penguins with small flippers but high body mass
penguins_clean_names |> 
  filter(flipper_length_mm < 185, body_mass_g > 4500) |> 
  select(species, sex, flipper_length_mm, body_mass_g, island)

```
:::{.callout-note}
Deviations from expected correlations might indicate:

Measurement error on one variable
Units confusion (one variable in wrong units)
Species misidentification
Legitimate individual variation (not always an error!)
:::

### Cross-variable checks: Biological impossibilities

Some variable combinations are biologically impossible:

```{r}
# Males cannot lay eggs
penguins_clean_names |> 
  filter(sex == "MALE", !is.na(date_egg)) |> 
  select(species, sex, date_egg, body_mass_g, island)

```

### Cross-variable checks: Spatial consistency

Location variables should match known species distributions:

```{r}
# Check which species appear on which islands
penguins_clean_names |> 
  count(species, island) |> 
  pivot_wider(names_from = island, values_from = n, values_fill = 0)

```

#### Known biology:

- Adelie penguins occur on all three islands

- Chinstrap penguins only on Dream island

- Gentoo penguins only on Biscoe island


### Flagging suspicious values

Rather than immediately removing suspicious values, best practice is to flag them for review:

```{r}
# Create flags for different types of potential issues
penguins_flagged <- penguins_clean_names |> 
  mutate(
    # Single-variable flags
    flag_impossible = case_when(
      body_mass_g <= 0 ~ "negative_or_zero_mass",
      flipper_length_mm <= 0 ~ "negative_or_zero_flipper",
      TRUE ~ NA_character_
    ),
    flag_implausible = case_when(
      body_mass_g < 2000 ~ "suspiciously_light",
      body_mass_g > 7000 ~ "suspiciously_heavy",
      TRUE ~ NA_character_
    ),
    
    # Cross-variable flags
    flag_species_size = case_when(
      species == "Adelie" & body_mass_g > 5000 ~ "Adelie_too_heavy",
      species == "Gentoo" & body_mass_g < 4000 ~ "Gentoo_too_light",
      TRUE ~ NA_character_
    ),
    # Any flag present?
    any_flag = !is.na(flag_impossible) | !is.na(flag_implausible) | 
               !is.na(flag_species_size) 
  )

# Summarize flagged observations
penguins_flagged |> 
  summarise(
    n_impossible = sum(!is.na(flag_impossible)),
    n_implausible = sum(!is.na(flag_implausible)),
    n_species_size = sum(!is.na(flag_species_size)),
    total_flagged = sum(any_flag)
  )

```
