
---
title: "What is Shiny? An Interactive Introduction"
author: "Your Name"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(ggplot2)
```

# 1. What is Shiny?

At its core, **Shiny is an R package that lets you build interactive web applications — using only R**. Instead of learning HTML, CSS, or JavaScript, you can create dashboards, data explorers, and even teaching tools right from your RStudio script.

Shiny is especially powerful when you're working with:
- 📊 **Interactive data visualizations**
- 🤝 **Sharing analysis with collaborators**
- 📣 **Communicating science in an accessible way**
- 🌉 **Bridging R users and non-R users through web UIs**

Below is a complete Shiny app that we will dissect and recreate during this workshop.

---

# 2. Try the App: Your First Shiny Demo

Run the full app below to preview where we’re headed:

```{r, echo=TRUE}
library(shiny)

ui <- fluidPage(
  titlePanel("Old Faithful Histogram"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
                  "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)
    ),
    mainPanel(
      plotOutput("distPlot")
    )
  )
)

server <- function(input, output) {
  output$distPlot <- renderPlot({
    x <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    hist(x, breaks = bins, col = 'skyblue', border = 'white')
  })
}

shinyApp(ui = ui, server = server)
```

### 💡 Did you know?
You don’t need to manually reload the app every time you make a change — just click “Run App” again and RStudio will update the view automatically.

---

# 3. Make It Your Own: Guided Modifications

## ✅ Challenge 1: Change the App Title
Change the `titlePanel()` to display your own custom app title. For example:

```r
titlePanel("My First Interactive App")
```

Re-run the app and confirm your new title appears.

---

## ✅ Challenge 2: Customize the Slider
Locate the `sliderInput()` function and update the label and default values:

```r
sliderInput("bins",
            "Choose the number of bins:",
            min = 5,
            max = 100,
            value = 20)
```

### 💡 What do these arguments control?
- `min`: Minimum slider value
- `max`: Maximum slider value
- `value`: Default setting

---

## ✅ Challenge 3: Add Description Text
Use the `p()` function to add an explanatory paragraph under the title:

```r
fluidPage(
  titlePanel("My First App"),
  p("This app shows a histogram of the waiting time between eruptions at Old Faithful geyser."),
  sidebarLayout(...)
)
```

---

## ✅ Challenge 4: Update the Plot Color
Inside `renderPlot()`, customize the histogram color:

```r
hist(x, breaks = bins, col = 'lightblue', border = 'grey30')
```

---

## ✅ Challenge 5: Use ggplot2 Instead
Replace `hist()` with `ggplot()` for better visuals:

```r
library(ggplot2)

output$distPlot <- renderPlot({
  ggplot(faithful, aes(waiting)) +
    geom_histogram(bins = input$bins,
                   fill = "steelblue3",
                   colour = "grey30") +
    xlab("Waiting time to next eruption") +
    theme_minimal()
})
```

---

## ✅ Challenge 6: Add a Radio Button to Select Variable
Let users choose between two variables:

```r
radioButtons(inputId = "display_var",
             label = "Which variable to display:",
             choices = c("Waiting time" = "waiting",
                         "Eruption time" = "eruptions"),
             selected = "waiting")
```

Add this to your `sidebarPanel()`.

### Then update your plot to respond:

```r
output$distPlot <- renderPlot({
  if (input$display_var == "eruptions") {
    xlabel <- "Eruption Time (min)"
  } else {
    xlabel <- "Waiting Time to Next Eruption (min)"
  }

  ggplot(faithful, aes(.data[[input$display_var]])) +
    geom_histogram(bins = input$bins,
                   fill = "steelblue3",
                   colour = "grey30") +
    xlab(xlabel) +
    theme_minimal()
})
```

---

## 🚀 Recap Challenge: Make It Yours

Customize your first Shiny app by doing the following:
- Use `ggplot2` to render the histogram
- Let the user choose the variable to plot
- Add a title and description
- Style the plot with your own colors and layout
